---
title: "MPulleyM2Q5"
author: "Melissa Pulley"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Download and read the orchid dataset that we used for the original logisticlab:

```{r}
orchid=readRDS("orchid.RDS")
orchid$habitat = as.factor(orchid$habitat)
summary(orchid)
```

Run the following models (assume Poisson distribution):

1. Null model

2. Effect of elevation on abundance

3. Effect of elevation and habitat on abundance (additive)

4. Interactive effect of elevation and habitat on abundance

5. Effect of habitat on abundance

```{r}
m1 <- glm(abundance ~ 1, family = poisson(link = "log"),  data = orchid)
m2 <- glm(abundance ~ elevation, family = poisson(link = "log"),  data = orchid)
m3 <- glm(abundance ~ elevation + habitat, family = poisson(link = "log"),  data = orchid)
m4 <- glm(abundance ~ elevation*habitat, family = poisson(link = "log"),  data = orchid)
m5 <- glm(abundance ~ habitat, family = poisson(link = "log"),  data = orchid)
```


Using AICc select the best model, interpret it, and plot it 

```{r}
library(MuMIn)
test=AICc(m1,m2,m3,m4,m5)
best = min(test$AICc)
test
best
```
Model 2 is the best model because it has the lowest AIC. The means that for abundance of orchids, elevation is an important explanatory variable.

```{r}
newdata_orchid = data.frame(abundance = seq(min(orchid$abundance), 
                                          max(orchid$abundance),length=100), 
                       elevation = seq(min(orchid$elevation), 
                                          max(orchid$elevation),length=100),
                       show.legend = TRUE)

pred.link.full1 <- predict(m2, newdata = newdata_orchid, se.fit = TRUE)
newdata_orchid$ab <- exp(pred.link.full1$fit) 

library(ggplot2)
ggplot() +
  geom_point(data = orchid, aes(x = elevation, y = abundance), size=2) +
  geom_path(data = newdata_orchid, aes(x = elevation, y = ab), linewidth=1) +
  xlab("Elevation") +
  ylab("Abundance") +
  labs(title="Impact of elevation on abundance of orchids")

```


Upload: AICc table, interpretation and plot.

Upload the code you used for the best model