---
title: "Week four examples"
author: "Alejandro Molina-Moctezuma"
date: "2024-02-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Population simulation

We will simulate a population. This simulated population has 5,438 mussels (Green Floater) in a stream. 
Size is represented by $ X \sim N(\mu = 40, \sigma = 6)$

```{r}
set.seed(123)
pop <- rnorm(n=5438, mean=40, sd=6)
hist(pop) 
```

## Empirical population mean size and sd

The following code obtains the **population parameters** for $\mu$ and $\sigma$. You can think of the estimated variables as **real values**

```{r}
mu <- mean(pop)
sigma <- sqrt(sum((pop - mu)^2)/length(pop)) ## why not just use sd()?
mu
sigma
```

## Let's sample!

We will do one snorkel survey. We snorkel and get 20 individuals.

```{r}
set.seed(NULL)
SnorkSample1 <- sample(pop, size=20)
SnorkSample1 ## see the sizes we have measured!
```



Look at your values. Those are the measurmenets of the 20 mussels you measured. Now, what's the mean and sd of your **sample** Sample estimates are our **approximations** to the real value by measureing a subsert of the population.


```{r}
mean(SnorkSample1)
sd(SnorkSample1)

```
This is for one sampling event! Now, what if we did it 1,000 times?

```{r}
Xbar <- rep(NA, 1000) # set up a vector of 1000 NAs

for (i in 1:1000){ 
	Xbar[i]<- mean(sample(pop, size=20))} 


mean(Xbar)
sd(Xbar)

```



And when we get a histogram?
```{r}
hist(Xbar, breaks=seq(min(Xbar),max(Xbar),length.out=26))
```




Setting a camera trap and see how many jaguars you see in a month


```{r}
 Xbar <- rep(NA, 40000) # set up a vector of 40000 NAs
 # calculate mean of a sample of 100 observations of Poisson random variable with # mean=4; repeat over 40000 times, i.e., take 40000 samples of Xbar
for (i in 1:40000){ 
     Xbar[i]<- mean(rpois(n=20, lambda=4))}
 hist(Xbar, breaks=seq(min(Xbar),max(Xbar),length.out=26))
#obtain histogram
```
