---
title: "Class examples 3"
author: "Alejandro Molina-Moctezuma"
date: "2024-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```
## Instructions

This is not a lab, but it will help you udnerstand the code needed to run 



## Case 1: Teporingo

The teporingos are an endangered species of rabbits. You are working at a global caprive breeding program, and you have been measuring the weight of each 1-yr old individual. Therefore, you know that your population mean is: 453.65g, and your population variance is 58.61315g.

You suspect one of the programs receiving Teporingos is mismanaged and not feeding them adequately, and you suspect they aren't gaining weight as they should.

You sample **20** teporingos, and their mass is **421.01 g**

My null hypothesis is:___

My alternative hypotesis is:___

### One tail test
We know that values between -1.96 and 1.96 contain 95% of the area under the curve.
We also know that everything to the left of -1.96 represents 2.5% of the area under the curve, but this time we are working on a on tail test.

What is the "Z" value that represent 5% of the area under the curve? Let's ask R!

```{r}
qnorm (.05)
```
Now, if our test-statistic is lower than that number, then, we can reject the null hypothesis!

The test statistic is estimated as:

$$Z = \displaystyle \frac{\bar{x}-\mu}{\sigma / \sqrt{n}}$$
Which results in:
```{r}

Zstat<-(421.01-453.65)/(58.61315/sqrt(20))
Zstat
```
Now, with that value, can we reject or fail to reject the null hypothesis?

Let's obtain the p-value
```{r}

pnorm(Zstat)
```
During the exam, you will be asked to interpret p values. Make sure you understand what's going on here!

Another way to estimate it directly is:
```{r}
pnorm(421.01, mean=453.65,sd=58.61315/sqrt(20))

```
Which gives us the same P-value. Understand where each number came from!

## t-test

### First step, import the data, and start the package

Download and import the .csv file using:

```{r}
teporingos<-read.csv("teporingos.csv")
teporingos<-teporingos$mass
```


Let's intstall a package. You only need to install a package once per computer
```{r}
#install.packages('lterdatasampler')
library(lterdatasampler)
```

Calculate the mean, and sample sd, as well as the n

```{r pressure, echo=FALSE}

X_bar<-
s<-
n<-
  
```

It's very easy to run a t-test in R.
We give R our data set, and tell it that we want to know if it's different from 453.65 (our population mean!)

```{r}

t.test(x=teporingos, mu=453.65, alternative='two.sided')

```

Now, there is an issue with that last line of code. Can you guess what it is?
Run it again, but do it ocrrectly this time
```{r}
t.test()

```

To estimate it by hand the equation is:

$$t = \displaystyle \frac{\bar{X}-\mu_all}{s / \sqrt{n}}$$

```{r}
t<-(X_bar-453.65)/(s/sqrt(n))
t
```

And we can estimate the p-value as well!

```{r}
pt(q=t,df=n-1) 
```

## Two sample unpaired test

Download the data:


```{r}
teporingos2<-read.csv("~teporingos2pops.csv")
teporingos2$site<-as.factor(teporingos2$site)
```

Let's explore the data:
```{r}
summary(teporingos2)
```

And the mean of both populations!

```{r}
tapply(teporingos2$mass,teporingos2$site,mean)
```
Now let's separate them by population

```{r}
teporingos_tex<- teporingos2[teporingos2$site %in% 'texcoco',]
teporingos_top<- teporingos2[teporingos2$site %in% 'topilejo',]
```

And let's run the t-test!
Let's assume equal variance:

```{r}
t.test(x=teporingos_tex$mass, y=teporingos_top$mass, alternative='two.sided', var.equal=T, paired=F)
```


And if we  had unequal variance:
```{r}
t.test(x=teporingos_tex$mass, y=teporingos_top$mass, alternative='two.sided', var.equal=F, paired=F)
```

That was easy!

## Paired test

```{r}
teporingospaired<-read.csv("~teporingospaired.csv")

```

Check the data:
```{r}
head(teporingospaired)
```

And run the data!

```{r}
t.test(teporingospaired$diet, teporingospaired$original, paired=T, alternative='two.sided')

```

Or, we can perform a one-sample t-test on the vector of differences (d), on whether its mean is significantly different from zero ## same answer!!

```{r}
teporingospaired$delta<-teporingospaired$diet - teporingospaired$original

t.test(teporingospaired$delta, mu=0, alternative='two.sided')
```

These tests were done using a two-tailed approach. Was that correct? Do it with one tail now!



## Non parametric tests

When we don't follow all assumptions. Mainly, normality

We can do a Mann-Whitney U test for unpaired data:

```{r}

forest <- c(7,8,15,9,11)
cropland <- c(4,6,8,3,9)
wilcox.test(forest, cropland, mu=0, alternative= 'two.sided')



```
This test essentially ranks each observation from lowest to highest, so the values aren't important, just the ranks:

```{r}
forestrank <- c(4,5.5,10,7.5,9); croplandrank <- c(2,3,5.5,1,7.5)
wilcox.test(forest, cropland, mu=0, alternative= 'two.sided') 

```

We can also do a "paired" test. Called Wilcoxon signed-rank test. We look at differences between pairs, rank those differences and look at the direction of the differences:

```{r}
forest <- c(7,8,15,9,11)
cropland <- c(4,6,8,3,9)
wilcox.test(forest, cropland, mu=0, paired=T, alternative= 'two.sided')
```

